---
title: "Exercise02 Multivariable model"
author: "Ian Handel"
date: "05/06/2019"
output: 
  html_document: 
    highlight: haddock
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = "")
```

This example builds on the previous one by adding a second predictor variable. 

In case they're not loaded already let's load the libraries we need and import the data. 

```{r, eval=FALSE}
library(tidyverse)
library(skimr)
dat <- read_csv("logreg_data_01_20190530.csv")
head(dat)
skim(dat)
```

```{r, include=FALSE}
library(tidyverse)
library(skimr)
dat <- read_csv("logreg_data_01_20190530.csv")
head(dat)
skim(dat)
```
In the previous exercise we predicted status from treatment Now we are going to also consider the animals age and see if it is associated with the disease status.

Our model with just treatment was this...

```{r}
mod1 <- glm(status=="diseased" ~ treatment,
           family = binomial,
           data = dat)

summary(mod1)
```

If we focus on the coefficients and turn them into odds ratios the odds of disease are about `r round(exp(coef(mod1))[[2]])` times greater for treated than un-treated controls

```{r}
exp(coef(mod1))
```

We can use `round(---, 3)` to round the numbers to 3 decimal places. 

```{r}
round(exp(coef(mod1)), 3)
```

Just as we did with linear regression we can add other predictors.

```{r}
mod2 <- glm(status=="diseased" ~ treatment + age,
           family = binomial,
           data = dat)

summary(mod2)
```

Again we can look at the odds ratios...

```{r}
round(exp(coef(mod2)), 3)
```

Note two things here...

First we now have an odds ratio for age. It's `r age_mult <- round(exp(coef(mod2))[[3]], 3)`. This means that for every day older an animal is it's odds of disease (all else being equal) increase by `r age_mult` times. So an animal 4 days older has odds `r age_mult` x `r age_mult` x `r age_mult` = `r round(age_mult^3, 3)` times greater of being diseased. Ask if this isn't clear!

Secondly the odds ratio of disease if treated have changed from `r round(exp(coef(mod1)), 3)[[2]]` to `r round(exp(coef(mod2)), 3)[[2]]`.

Why do you think this has happened?




